---
title: "corr EDA options"
author: " "
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  html_document:
    toc: false
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning=FALSE)

library(tidyverse)
library(DALEX)  #to get fifa data
library(ggpubr) #to show r values

library(cluster)
library(ggdendro)

fifa_df <- fifa %>% as_tibble() %>% 
  slice_sample(n = 500) %>% 
  mutate(BMI = weight_kg/(height_cm/100)^2, 
         value_eur = log(value_eur)) %>% 
  filter(goalkeeping_positioning <50) # remove goalkeepers 


glimpse(fifa_df)

fifa_df %>% 
  ggplot(aes(value_eur))+
  geom_histogram()

```


## Exploring Correlations {.tabset}

I've been dealing with some fairly wide data sets (70-100 columns), and trying to get a sense of the correlations between so many variables and my target has been challenging. In looking for some alternatives to a really big table or heatmap, I've put together these three options using a small portion of the FIFA data set. 

I'm sure there are better and cleaner ways to do these things, feel free to suggest alternatives!

### Grouped scatterplots

Starting with a simple scatter plot, select the outcome variable and any comparisons you want to see. This is a cleaner way to combine sets of variables and into more easily digestible chunks.

```{r}
fifa_df %>% 
  select(value_eur, contains("power_")) %>% 
  pivot_longer(-value_eur) %>% 
  ggplot(aes(value, value_eur))+
  geom_point(alpha = .2, color = "midnightblue")+
  geom_smooth(method = "lm", se = F, color = "black", lty = 1)+
  stat_cor(p.accuracy = 0.001, r.accuracy = 0.01, cor.coef.name = 'r', aes(label = paste(..r.label.., ..p.label.., sep = "~`,`~")))+
  facet_wrap(~name, scales = "free")+
  labs(x = NULL, y = "Value", title = "Power factors") +
  theme_minimal()+
  theme(strip.background = element_rect(fill = "grey40", color = "grey80", size = 1),
        strip.text = element_text(colour = "white"),
        plot.title = element_text(hjust = 0.5, face = "bold"))


fifa_df %>% select(value_eur, contains("mentality_")) %>% 
  pivot_longer(-value_eur) %>%
  ggplot(aes(value, value_eur))+
  geom_point(alpha = .7, color = "lightblue")+
  geom_smooth(method = "lm", se = F, color = "black", lty = 1)+
  stat_cor(p.accuracy = 0.001, r.accuracy = 0.01, cor.coef.name = 'r', aes(label = paste(..r.label.., ..p.label.., sep = "~`,`~")))+
  facet_wrap(~name, scales = "free")+
  labs(x = NULL, y = "Value", title = "Mentality factors") +
  theme_minimal()+
  theme(strip.background = element_rect(fill = "grey40", color = "grey80", size = 1),
        strip.text = element_text(colour = "white"),
        plot.title = element_text(hjust = 0.5, face = "bold"))



```

### Barplot of all correlations

This is a way to see the correlation coefficients, with colored bars indicating statistical significance. 

```{r}

fifa_corrs <- fifa_df %>% select(where(is.numeric)) %>% rstatix::cor_mat()  #correlation matrix of all numeric variables
fifa_corrs_p <-fifa_corrs %>% rstatix::cor_get_pval()    # p value for all correlatons

fifa_corrs_bind <- fifa_corrs %>% filter(rowname == "value_eur") %>%   # select just the row of the target variable
  bind_rows(fifa_corrs_p %>% filter(rowname == "value_eur"))

# rename the two rows
fifa_corrs_bind[1,1] <- "corr"
fifa_corrs_bind[2,1] <- "p_val"


# transposing the table 
fifa_corrs_bind_filtered <- fifa_corrs_bind %>% 
  pivot_longer(-rowname) %>% 
  filter(rowname == "corr") %>% 
  rename(r_val = value,
         measure = rowname)

fifa_pval_bind_filtered <- fifa_corrs_bind %>% 
  pivot_longer(-rowname) %>% 
  filter(rowname == "p_val") %>% 
  rename(p_val = value)

fifa_corrs_joined <- fifa_corrs_bind_filtered %>% 
  left_join(fifa_pval_bind_filtered, by = "name") %>% 
  filter(name != "value_eur")

fifa_cor_colors <- c("gray 41", "limegreen")   #pick two colors

fifa_corrs_joined %>% 
  ggplot(aes(fct_reorder(name, r_val), r_val, fill = p_val < 0.05))+   #can adjust for multiple comparisons if desired
  geom_col()+
  coord_flip()+
  scale_fill_manual(values = fifa_cor_colors)+
  labs(x = NULL, y = NULL, title = "Pearson correlation with player value", subtitle = "Colored bars indicate p < 0.05") +
  theme_minimal()+
  theme(legend.position = "none")

```

### Dendrogram

I saw a [really cool paper](https://arxiv.org/pdf/2105.06811.pdf){target="_blank"} that used agglomerative clustering of features by their correlations to produce a dendrogram. This is seemed like a really creative way to show a lot of variables. 

There are several ways to make a dendrogram in r. I wanted to color the variables according to their clusters, which is easy using {dendextend}, but I also preferred to use ggplot. For that we can use {ggdendro}, but there wasn't an easy way to color according to subgroups. This should be simple enough, but I struggled to find a simple way so I had to get a bit creative.

```{r include=FALSE}

fifa_corr_tbl <- fifa_df %>% select(where(is.numeric))  %>%
  corrr::correlate() %>% 
  mutate(across(everything(), ~ ifelse(is.na(.), 1, .))) %>% 
  select(-term)

fifa_corr_names <- fifa_df %>% select(where(is.numeric))  %>%
  corrr::correlate() %>% 
  mutate(across(everything(), ~ ifelse(is.na(.), 1, .))) %>% 
  select(term) %>% 
  mutate(rownames_to_column(.))


# hierarchal clustering

#define linkage methods
m <- c( "average", "single", "complete", "ward")
names(m) <- c( "average", "single", "complete", "ward")


#function to compute agglomerative coefficient
ac <- function(x) {
  agnes(fifa_corr_tbl, method = x)$ac
}

#calculate agglomerative coefficient for each clustering linkage method
purrr::map_dbl(m, ac)

#perform hierarchical clustering using Ward's minimum variance (highest from above)
clust <- agnes(fifa_corr_tbl, method = "ward")


dend_data <- dendro_data(clust)
dend_segments <- dend_data$segments

cuts <- dendextend::cutree(clust, k = 11) %>% as_tibble() %>% rownames_to_column()

dend_label_new <- dend_data$labels %>%  as_tibble() %>% 
  left_join(fifa_corr_names, by = c("label" = "rowname")) %>% 
  left_join(cuts, by = c("label" = "rowname")) %>% 
  mutate(value = as.factor(value))

colors11 <- c("#B57F50", "#33658A", "#7A306C", "#59C3C3", "#BD93BD", "#55C1FF", "#BDB2FF", "#5DA9E9", "#588157",
              "#B57F51", "#33668A")



```

```{r fig.height=8, fig.width=8}

dend_segments %>% 
  ggplot() + 
  geom_segment(aes(x = x, y = y, xend = xend, yend = yend))+
  geom_text(data = dend_label_new, aes(x, y, label = term, color = value),
            hjust = 1, vjust = 0, angle = 0, size = 3, nudge_y = -0.1)+
  ylim(-5, 23) +
  #geom_hline(yintercept = 2)+
  coord_flip() +
  labs(x = NULL, y = NULL, title = "Dendrogram of hierarchical correlational clustering") +
  scale_color_manual(values = colors11) +
  theme(
        legend.position = "none",
        plot.title = element_text(hjust = 0.5, face = "bold"))+
  theme_dendro()

```

